Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-31 parallel.tex
Fixing formatting issues with underscores in texttt.
Added a table of variables and constants used in the paper.
Minor revision to parallelism section.

The current version of \enzo\ has been parallelized for distributed
memory platforms using the Message Passing Interface (MPI).  This is
done using a single grid object as the basic unit of parallelization.
Each grid object -- including all cell and grid data -- is fully
contained in a single processor\footnote{In this context, we use the
word {\it processor} to mean a basic distribution unit; this could by
either a core or a node, depending on details of the system.  Note
that the current version of Enzo is not threaded, although the
development of a hybrid version is in process}.  Parallelization is
accomplished by distributing grids amongst processors.  This is done
on the root grid using a simple tiling system, where the root grid is
split up into $N_{\rm root}$ tiles, with $N_{\rm root}$ typically
equal to the number of processors $N_p$.

@@

Because of the wide variety of simulations, methods, and uses of Enzo,
it is difficult to state in general terms which routines within the
code will be most costly during a given simulation.  As such, we have
designed a lightweight registering system that has been implemented
for the most commonly used routines (such as the hydrodynamic and
gravity solvers) as well as refinement level timers that measure the
time spent on each level.  Beyond this minimal set of routines, we
have designed a simple way for the user to modify the source by adding
\texttt{TIMER\_START(``Your_Routine_Name'')} and
\texttt{TIMER\_END(``Your_Routine_Name'')}.  These timers are created
and managed individually on each processor in an asynchronous fashion,
and contribute minimal computational, memory, and IO overhead.

At each complete root grid time step (or less often if specified),
each timer is then communicated to the root processor where it
calculates the mean, standard deviation, minimum, and maximum for each
of the timers across all processors.  For level timers, there are additional
attributes such as the number of cell updates, the current number of
grids, and the average cells/s/MPI process.  This information is then
output to a logfile.  This provides a simplified interface to the user
that can be used to diagnose performance issues as well as estimate a
given problem type's scalability.  In addition to the logfile, we have
developed a plotting interface for quickly producing figures that
process the data from the logfile.  These capabilities are described
in the online documentation, along with a further discussion of the
performance measurement implementation.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/enzo_unigrid_weak_scaling.eps}
\caption{\enzo\ weak scaling performance for a set of Lyman alpha
forest cosmology simulations with constant comoving spatial resolution
per grid cell, showing cell updates per second per computational core
plotted as a function of the number of root grid tiles of dimension
$128^3$ (R) in each dimension.  The number of MPI tasks is N$ = R^3$,
so R$ = 16$ on this plot corresponds to a $2048^3$ computational mesh
running on 4,096 MPI tasks.  This plot goes from R$ = 2$ (8 MPI tasks)
to R$ = 24$ (13,824 MPI tasks) on two supercomputers -- NICS Kraken
and ORNL Jaguar when they were Cray XT4 systems -- and using 1, 2 or 4
MPI processes per node, where each compute node contained a single
quad-core AMD Opteron CPU having a speed of 2.1 GHz on Jaguar and 2.3
GHz on Kraken.}
\label{fig.uniscale}
\end{center}
\end{figure}

It is advantageous to use \enzo\ in its ``unigrid'' (i.e.,
non-adaptive mesh) mode for a variety of problems, including
non-gravitating turbulence
\citep[e.g.,][]{2002ApJ...569L.127K,Kritsuk04}, the Lyman-alpha forest
\citep{2005MNRAS.361...70J,2009MNRAS.399.1934P}, or feedback of
metal-enriched gas from galaxies into the intergalactic medium
\citep{2004ApJ...601L.115N,2011ApJ...731....6S}.  Achieving good
scaling of the code in unigrid mode is relatively straightforward --
upon initialization, unigrid simulations are decomposed such that each
MPI process has a roughly equal subvolume (and thus number of grid
cells), meaning that work is evenly distributed.  Communication
patterns for both the gravity solve (which uses a fast Fourier
transform) and the fluid solves (which transfer boundary information
between subvolumes) are predictable and straightforward, and
rebuilding of the grid hierarchy does not take place, removing a
substantial global operation and a great deal of communication.

Figure~\ref{fig.uniscale} shows \enzo\ weak scaling results for a
sequence of scaled unigrid Lyman alpha forest calculations. These
calculations include dark matter dynamics, hydrodynamics using the
piecewise parabolic method, six-species non-equilibrium chemistry and
radiative cooling, and a uniform metagalactic ultraviolet background.
In this sequence of test calculations, we perform a weak scaling test
on up to 13,824 MPI tasks on the NICS Kraken XT4 and ORNL Jaguar XT4
supercomputers\footnote{These simulations were performed prior to
conversion of both machines to the current-generation systems}.  In
this test, each MPI task was given a $128^3$ root grid tile (i.e.,
$128^3$ grid cells containing baryon quantities) and, initially,
approximately $128^3$ dark matter particles.  The number of grid cells
was constant throughout the calculation; the number of dark matter
particles varies as they are moved from subvolume to subvolume as
structure evolves.  The grid resolution was kept at a constant
comoving size of $\simeq 40$~kpc/h, and as the core count was
increased, so was the simulation volume.  On each machine, a compute
node contained a single AMD Opteron quad-core chip (2.1 Ghz on Jaguar;
2.3 Ghz on Kraken) with 2 GB/memory per core (8 GB/total per node).
Both machines used the SeaStar2 interconnect.  In the scaling study,
calculations were run with 1, 2, or 4 MPI tasks per node.  The figure
shows cell updates per second per MPI process; perfect scaling would
be a horizontal line.

As can be seen in Figure~\ref{fig.uniscale}, the unigrid weak scaling
performance of the code is extremely good for this problem, with only
a 20\% decrease in cell updates/second/task as the code is scaled from
8 to 4,096 MPI tasks, and a 40\% decrease in performance overall going
from 8 to 13,824 (or $24^3$) MPI tasks.  We speculate that this
decrease is likely to be partially due to global MPI communications
used to, e.g., calculate the overall timestep of the simulation, and
also likely due to load imbalances due to increasing cosmological
power (and thus an increasingly uneven distribution of dark matter
particles between MPI tasks at late times) as the simulation volume
grows.  We also observe that a systematic difference in speed can be
seen between the two machines, which can be attributed primarily to
the slightly faster CPUs on Kraken at th etime (2.3 Ghz, vs. 2.1 Ghz
on Jaguar).  The difference in speed when using different numbers of
MPI tasks per node can be attributed primarily due to differences in
competing usage of shared cache on the quad-core chips used on this
machine.

@@

Many astrophysical problems, such as cosmological galaxy formation
\citep{2012ApJ...749..140H}, high-resolution disk galaxy simulations
\citep{2011ApJ...738...54K}, high-redshift \citep{2009Sci...325..601T}, and present-day
star formation \citep{Collins12a}, involve multi-scale physics that
span several orders of magnitude in both space and time.  In these
situations, using Enzo in its adaptive mesh refinement mode is
beneficial.  Because the adaptive grid hierarchy is dynamic, grid
boundaries and, thus, communication patterns can be unpredictable,
hindering strong scaling behavior to high core counts.

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\textwidth]{figures/scaling_plot2}}
\caption{This figure show (clockwise from upper left), the covering
fraction $f_l$ of the grids on a given level $l$, the ratio of
$f_{l}/f_{l+1}$, the relative memory usage of each level (normalized
by the memory usage of level 0), and, in the lower-left panel, the
relative CPU usage of each level, again normalized by $l=0$.  In each
panel, three curves are shown: the solid black and long dashed red are
two cosmological simulations with box sizes of 30 h$^{-1}$ Mpc and
80h$^{-1}$ Mpc respectively.  The dot-dashed blue line is for a
non-cosmological simulation of a ram-pressure stripped galaxy.  In the
top-right panel, there is a line at the critical ratio of 0.06, which
determines if the next finer level ($l+1)$ takes more (above) or less
(below) CPU time than level $l$.}
\label{fig:scaling}
\end{figure}

As, the figure demonstrates, the $f_l$ values all show a sharp decline
with level, dropping as power-laws with the level $l$.  Focusing first
on the ``refine-everywhere" cosmological runs, which both show similar
behavior despite the different box sizes and redshifts at which we
collect the data.  In fact, we find these results are very typical for
cosmological runs, with only slight variations depending on the exact
system that is being simulated.  The top-left panel
shows the ratio of $f_{l}/f_{l+1}$ which, remarkably, hovers around
the critical value of 0.0625 determined earlier.  This implies that
the relative CPU usage of each level, shown in the lower-left panel,
is nearly flat, and that going to additional levels of refinement only
increases the amount of calculation by a factor of roughly $1/l_{\rm
max}$.  Even in the most extreme case, the L80 run, the last level
adds only 25\% to the computation time.

The memory used by each level, on the other hand, scales as
\begin{equation}
{\rm memory} = C_2 \sum_l f_l N_{\rm root}^3 2^{3l}
\end{equation}
and the terms in this sum are shown in the bottom-right panel of
Figure~\ref{fig:scaling}.  This demonstrates that the memory usage is
dominated by the top three levels, and adding additional levels only
adds minimally to the memory usage of the run.  We have empirically
tested this scaling for cosmological simulations and found it to be
reasonably accurate (although the increase is typically slightly
larger than found here, usually 30-50\%, probably because of
less-than-ideal parallel scaling).

>>>>

The current version of \enzo\ has been parallelized for distributed
memory platforms using the Message Passing Interface (MPI).  This is
done using a single grid object as the basic unit of parallelization.
Each grid object -- including all cell and grid data -- is fully
contained on a single processor\footnote{In this context, we use the
word {\it processor} to mean a basic distribution unit; this could be
either a core or a node, depending on details of the system.  Note
that the current version of Enzo is not threaded, although a hybrid
MPI + OpenMP version is currently under development.}.  
Parallelization is
accomplished by distributing grids amongst processors.  This is done
on the root grid using a simple tiling system, where the root grid is
split up into $N_{\rm root}$ tiles, with $N_{\rm root}$ typically
equal to the number of processors, $N_p$.

@@

Because of the wide variety of simulations, methods, and uses of Enzo,
it is difficult to state in general terms which routines within the
code will be most costly during a given simulation.  As such, we have
designed a lightweight registering system that has been implemented
for the most commonly used routines (such as the hydrodynamic and
gravity solvers) as well as refinement level timers that measure the
time spent on each level.  Beyond this minimal set of routines, we
have designed a simple way for the user to modify the source by adding
\texttt{TIMER\_START("Your\_Routine\_Name")} and
\texttt{TIMER\_END("Your\_Routine\_Name")}.  These timers are created
and managed individually on each processor in an asynchronous fashion,
and contribute minimal computational, memory, and IO overhead.

At each complete root grid time step (or less often if specified),
each timer is then communicated to the root processor where it
calculates the mean, standard deviation, minimum, and maximum for each
of the timers across all processors.  For level timers, there are additional
attributes such as the number of cell updates, the current number of
grids, and the average cells updates per second per MPI process.  This
information is then 
output to a logfile.  This provides a simplified interface to the user
that can be used to diagnose performance issues as well as estimate a
given problem type's scalability.  In addition to the logfile, we have
developed a plotting interface for quickly producing figures that
process the data from the logfile.  These capabilities are described
in the online documentation, along with further discussion of the
performance measurement implementation.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/enzo_unigrid_weak_scaling.eps}
\caption{\enzo\ weak scaling performance for a set of Lyman-$\alpha$
forest cosmology simulations with constant comoving spatial resolution
per grid cell, showing cell updates per second per processor 
plotted as a function of the number of root grid tiles of dimension
$128^3$ (R) in each dimension.  The number of MPI tasks is N$ = R^3$,
so R$ = 16$ on this plot corresponds to a $2048^3$ computational mesh
running on 4,096 MPI tasks.  This plot goes from R$ = 2$ (8 MPI tasks)
to R$ = 24$ (13,824 MPI tasks) on two supercomputers -- NICS Kraken
and ORNL Jaguar when they were Cray XT4 systems -- and using 1, 2 or 4
MPI processes per node, where each compute node contained a single
quad-core AMD Opteron CPU with a speed of 2.1 GHz on Jaguar and 2.3
GHz on Kraken.}
\label{fig.uniscale}
\end{center}
\end{figure}

It is advantageous to use \enzo\ in its ``unigrid'' (i.e.,
non-adaptive mesh) mode for a variety of problems, including
non-gravitating turbulence
\citep[e.g.,][]{2002ApJ...569L.127K,Kritsuk04}, the Lyman-$\alpha$ forest
\citep{2005MNRAS.361...70J,2009MNRAS.399.1934P}, or feedback of
metal-enriched gas from galaxies into the intergalactic medium
\citep{2004ApJ...601L.115N,2011ApJ...731....6S}.  Achieving good
scaling of the code in unigrid mode is relatively straightforward --
upon initialization, unigrid simulations are decomposed such that each
MPI process has a roughly equal subvolume (and thus an equal number of grid
cells), meaning that work is evenly distributed.  Communication
patterns for both the gravity solve (which uses a fast Fourier
transform) and the fluid solves (which transfer boundary information
between subvolumes) are predictable and straightforward, and
rebuilding of the grid hierarchy does not take place, removing a
substantial global operation and a great deal of communication.

Figure~\ref{fig.uniscale} shows \enzo\ weak scaling results for a
sequence of scaled unigrid Lyman-$\alpha$ forest calculations. These
calculations include dark matter dynamics, hydrodynamics using the
PPM solver, six-species non-equilibrium chemistry and
radiative cooling, and a uniform metagalactic ultraviolet background.
In this sequence of test calculations, we perform a weak scaling test
on up to 13,824 MPI tasks on the NICS Kraken XT4 and ORNL Jaguar XT4
supercomputers\footnote{These simulations were performed prior to
conversion of both machines to the current-generation systems}.  In
this test, each MPI task was given a $128^3$ root grid tile (i.e.,
$128^3$ grid cells containing baryon quantities) and, initially,
approximately $128^3$ dark matter particles.  The number of grid cells
on each processor 
was constant throughout the calculation; the number of dark matter
particles on each processor varies as they are moved from subvolume to subvolume as
structure evolves.  The grid resolution was kept at a constant
comoving size of $\simeq 40$~kpc/h, and as the core count was
increased, so was the simulation volume.  On each machine, a compute
node contained a single AMD Opteron quad-core chip (2.1 Ghz on Jaguar;
2.3 Ghz on Kraken) with 2 GB/memory per core (8 GB/total per node).
Both machines used the SeaStar2 interconnect.  In the scaling study,
calculations were run with 1, 2, or 4 MPI tasks per node.  The figure
shows cell updates per second per MPI process; perfect scaling would
be a horizontal line.

As can be seen in Figure~\ref{fig.uniscale}, the unigrid weak scaling
performance of the code is extremely good for this problem, with only
a 20\% decrease in cell updates per second per MPI task as the code is scaled from
8 to 4,096 MPI tasks, and a 40\% decrease in performance overall going
from 8 to 13,824 (or $24^3$) MPI tasks.  We speculate that this
decrease is likely to be partially due to global MPI communications
used to, e.g., calculate the overall timestep of the simulation, and
also likely due to load imbalances due to increasing cosmological
power (and thus an increasingly uneven distribution of dark matter
particles between MPI tasks at late times) as the simulation volume
grows.  We also observe that a systematic difference in speed can be
seen between the two machines, which can be attributed primarily to
the slightly faster CPUs on Kraken at th etime (2.3 Ghz, vs. 2.1 Ghz
on Jaguar).  The difference in speed when using different numbers of
MPI tasks per node can be attributed primarily due to differences in
competing usage of shared cache on the quad-core chips used on this
machine.

@@

Many astrophysical problems, such as cosmological galaxy formation
\citep{2012ApJ...749..140H}, high-resolution disk galaxy simulations
\citep{2011ApJ...738...54K}, high-redshift \citep{2009Sci...325..601T}, and present-day
star formation \citep{Collins12a}, involve multi-scale physics that
span several orders of magnitude in both space and time.  In these
situations, using Enzo in its adaptive mesh refinement mode is
beneficial.  Because the adaptive grid hierarchy is dynamic, grid
boundaries and, thus, communication patterns can be unpredictable,
hindering strong scaling to high core counts.

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\textwidth]{figures/scaling_plot2}}
\caption{This figure shows (clockwise from upper left), the covering
fraction $f_l$ of the grids on a given level $l$, the ratio of
$f_{l}/f_{l+1}$, the relative memory usage of each level (normalized
by the memory usage of level 0), and, in the lower-left panel, the
relative CPU usage of each level, again normalized by $l=0$.  In each
panel, three curves are shown: the solid black and long dashed red are
two cosmological simulations with box sizes of 30 h$^{-1}$ Mpc and
80h$^{-1}$ Mpc respectively.  The dot-dashed blue line is for a
non-cosmological simulation of a ram-pressure stripped galaxy.  In the
top-right panel, there is a line at the critical ratio of 0.06, which
determines if the next finer level ($l+1)$ takes more (above) or less
(below) CPU time than level $l$.}
\label{fig:scaling}
\end{figure}

As, the figure demonstrates, the $f_l$ values all show a sharp decline
with level, dropping as power-laws with the level $l$.  Focusing first
on the ``refine-everywhere" cosmological runs, which both show similar
behavior despite the different box sizes and redshifts at which we
collect the data.  In fact, we find these results are very typical for
cosmological runs, with only slight variations depending on the exact
problem that is being simulated.  The top-left panel
shows the ratio of $f_{l}/f_{l+1}$ which, remarkably, hovers around
the critical value of 0.0625 determined earlier.  This implies that
the relative CPU usage of each level, shown in the lower-left panel,
is nearly flat, and that going to additional levels of refinement only
increases the amount of calculation by a factor of roughly $1/l_{\rm
max}$.  Even in the most extreme case, the L80 run, the last level
adds only 25\% to the computation time.

The memory used by each level, on the other hand, scales as
\begin{equation}
{\rm memory} = C_2 \sum_l f_l N_{\rm root}^3 2^{3l}
\end{equation}
and the terms in this sum are shown in the bottom-right panel of
Figure~\ref{fig:scaling}.  This demonstrates that the memory usage is
dominated by the top three levels, and adding additional levels 
adds minimally to the memory usage of the run.  We have empirically
tested this scaling for cosmological simulations and found it to be
reasonably accurate (although the increase is typically slightly
larger than found here, usually 30-50\%, probably because of
less-than-ideal parallel scaling).
